---
title: April Timing Forecast
author: Bryce Mecum
output: md_document
---

# April Timing Forecast

```{r packages, include=FALSE}
suppressPackageStartupMessages({
  library(dplyr)
  library(ggplot2)
  library(readr)
})
```

```{r load_data, include=FALSE}
yuk <- read_csv("../data/yukon.csv")
amatc <- yuk[yuk$year == 2018, "amatc"]
```

## Data Aquisition

The 2018 value of AMATC was retrieved using manually using http://w2.weather.gov/climate/xmacis.php?wfo=pafg. The value was -3.2 °C.

## Graphical Analysis

How does the 2018 value of AMATC, `r amatc`, compare to the historical data?

```{r amatc_plot, echo=FALSE, warning=FALSE, fig.height=4, fig.width=4}
ggplot(yuk, aes(amatc, mdj)) +
  geom_point(shape = 1) +
  geom_vline(xintercept = as.numeric(yuk[which(yuk$year == 2018),"amatc"])) +
  labs(x = expression("AMATC,"*~degree*"C"), y = "Median Run Timing (June)")
ggsave("mdj_against_amatc.png", width = 4, height = 4)
```

**Intepretation:** An AMATC of -3.2 °C slightly warmer than average (`r round(mean(yuk$amatc), 1)` °C) so we would expect a slightly early run timing.

## Forecast

```{r, echo=FALSE, include=FALSE}
model <- lm(mdj ~ amatc, data = subset(yuk, year < 2018))

predicted <- floor(unname(predict(model, newdata =  yuk[yuk$year == 2018,])))
```

To forecast MDJ for 2018, I followed the approach as in previous years:

I fit a simple linear model of AMATC vs. MDJ using 55 years (1961 -- 2017) of historical AMATC and MDJ.
I then predicted the 2018 MDJ using the fitted model which came out to be **June `r predicted`** which is just slightly earlier than average (mean `r floor(mean(yuk$mdj, na.rm = TRUE))`).

### Hindcasting

```{r, echo=FALSE}
result <- data.frame()

for (y in (max(yuk$year) - 15):(max(yuk$year) - 1)) {
  model <- lm(mdj ~ amatc, data = subset(yuk, year < y))
  model_pred <- as.numeric(floor(predict(model, newdata =  yuk[yuk$year == y,"amatc"])))
  
  result <- rbind(result,
                  data.frame(year = y,
                             obs = as.numeric(yuk[yuk$year == y,"mdj"]),
                             pred = round(model_pred, 0)))
}


metric_mad <- round(mean(abs(result$obs - result$pred)), 2)
metric_mad_sd <- round(sd(abs(result$obs - result$pred)), 2)
metric_max_abs_resid <- round(max(abs(result$obs - result$pred)), 2)
# mean(result$obs - result$pred)
```

As with the full forecast and previous years, I used a hindcasting approach (one-step-ahead cross-validation) to get a sense of how well AMATC predicts MDJ.

I hindcasted MDJ using an arbitrary window of 2003 -- 2017 (n=15) using the same AMATC vs. MDJ model as above and calculated the following three metrics:

| Metric | Value (days) |
|--------|-------|
| Mean absolute devitation | `r metric_mad` |
| Standard deviation of absolute deviations |`r metric_mad_sd`|
| Maximum absolute residual | `r metric_max_abs_resid` |
| Mean deviation | `r round(mean(result$pred - result$obs))`

Predicted values of MDJ were rounded down to the nearest day because MDJ is recorded at daily time steps.

**Interpretation**: The AMATC-only model is a lot less useful than the full model and tends to generate predictions that are four days off the true MDJ and tend to be biased 4 days early.